Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-31 draft_lm1.tex
important reference

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes, an intrinsically stochastic events, can drastically change the 
pool of susceptibles, throwing off any sequential estimation based on the 
incidence dynamics.

@@

\subsection*{Data} 
The data used in this paper consists of time 
series of weekly notified cases of dengue for the 
city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical criteria. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalization was done by dividing the number of cases reported 
by the total city population at each year as given by the census.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R convergence diagnostic~\cite{brooks1998}.

>>>>

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
Recently, methods in the same fashion were developed for dengue when 
serotype-specific data is available~\cite{Reiner2014}.
When such data is not available, the series of susceptibles to all possible
serotypes, cannot be reconstructed based solely on a deterministic transmission 
model, since the arrival/re-emergence of  new serotypes, an intrinsically 
stochastic events, can drastically change the pool of susceptibles, throwing off 
any sequential estimation based on the  incidence dynamics.

@@

\subsection*{Data}

The data used in this paper consists of time 
series of weekly notified cases of dengue for the 
city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical criteria.
Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013).
For the parameter estimation  procedures incidence was normalized by dividing 
the number of cases reported by the total city population at each year as given 
by the census (Census Bureau, Brazilian Institute of Geography and Statistics, 
\url{http://www.ibge.gov.br/english/}).

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R (potential scale 
reduction), that approaches $1$ at convergence~\cite{brooks1998}.
